Memory for television episodes preserves event content while introducing new across-event similarities

### 2020-02-07 paper/main.tex
finished initial revision of paper

\subsubsection*{Visualizing the video and recall topic trajectories}
We used the UMAP algorithm~\citep{McInEtal18} to project the 100-dimensional topic space onto a two-dimensional space for visualization (Figs.~\ref{fig:trajectory}, \ref{fig:topics}).  Importantly, to ensure that all of the trajectories were projected onto the \textit{same} lower dimensional space, we computed the low-dimensional embedding on a ``stacked'' matrix created by vertically concatenating the events-by-topics topic proportions matrices for the video, across-participants average recalls and all 17 individual participants' recalls.  We then divided the rows of the result (a total-number-of-events by two matrix) back into separate matrices for the video topic trajectory and the trajectories for each participant's recalls (Fig.~\ref{fig:trajectory}).  This general approach for discovering a shared low-dimensional embedding for a collections of high-dimensional observations follows \cite{HeusEtal18a}.  Note: for further details on how we created this low-dimensional embedding space, see \textit{Supporting Information}.
% ------ NOTE: ------
% need to add info about embedding optimization
% - Paxton

\subsubsection*{Estimating the consistency of flow through topic space across participants}
In Figure~\ref{fig:trajectory}B, we present an analysis aimed at characterizing locations in topic space that different participants move through in a consistent way (via their recall topic trajectories).  The two-dimensional topic space used in our visualizations (Fig.~\ref{fig:trajectory}) comprised a 60 x 60 (arbitrary units) square.  We tiled this space with a 50 x 50 grid of evenly spaced vertices, and defined a circular area centered on each vertex whose radius was two times the distance between adjacent vertices (i.e., 2.4 units).  For each vertex, we examined the set of line segments formed by connecting each pair successively recalled events, across all participants, that passed through this circle.  We computed the distribution of angles formed by those segments and the $x$-axis, and used a Rayleigh test to determine whether the distribution of angles was reliably ``peaked'' (i.e., consistent across all transitions that passed through that local portion of topic space).  To create Figure~\ref{fig:trajectory}B we drew an arrow originating from each grid vertex, pointing in the direction of the average angle formed by line segments that passed within its circular radius.  We set the arrow lengths to be inversely proportional to the $p$-values of the Rayleigh tests at each vertex.  Specifically, for each vertex we converted all of the angles of segments that passed within 2.4 units to unit vectors, and we set the arrow lengths at each vertex proportional to the length of the (circular) mean vector.  We also indicated any significant results ($p < 0.05$, corrected using the Benjamani-Hochberg procedure) by coloring the arrows in blue (darker blue denotes a lower $p$-value, i.e., a longer mean vector); all tests with $p \geq 0.05$ are displayed in gray and given a lower opacity value.

\subsection*{Searchlight fMRI analyses}
In Figure~\ref{fig:brainz}, we present two analyses aimed at identifying brain regions whose responses (as participants viewed the video) exhibited a particular temporal structure.  We developed a searchlight analysis wherein we constructed a cube centered on each voxel (radius: 5~voxels) and for each of these cubes, computed the temporal correlation matrix of the voxel responses during video viewing.  Specifically, for each of the 1976 volumes collected during video viewing, we correlated the activity patterns in the given cube with the activity patterns (in the same cube) collected during every other timepoint.  This yielded a 1976 by 1976 correlation matrix for each cube.

Next, we constructed a series of ``template'' matrices: the first reflecting the timecourse of video's topic trajectory, and the others reflecting that of each participant's recall topic trajectory.  To construct the video template, we computed the correlations between the topic proportions estimated for every pair of TRs (prior to segmenting the trajectory into discrete events; i.e., the correlation matrix shown in Figs.~\ref{fig:model}B and \ref{fig:brainz}A).  We constructed similar temporal correlation matrices for each participant's recall topic trajectory (Figs.~\ref{fig:model}D, \corrmats).  However, to correct for length differences and potential non-linear transformations between viewing time and recall time, we first used dynamic time warping~\citep{BernClif94} to temporally align participants' recall topic trajectories with the video topic trajectory.  An example correlation matrix before and after warping is shown in Fig.~\ref{fig:brainz}B.  This yielded a 1976 by 1976 correlation matrix for the video template and for each participant's recall template.

To determine which (cubes of) voxel responses matched the video template, we correlated the upper triangle of the voxel correlation matrix for each cube with the upper triangle of the video template matrix~\citep{KrieEtal08b}.  This yielded, for each participant, a voxelwise map of correlation values.  We then performed a one-sample $t$-test on the distribution of (Fisher $z$-transformed) correlations at each voxel, across participants.  This resulted in a value for each voxel (cube), describing how reliably its timecourse mirrored that of the video.

We further sought to ensure that our analysis identified regions where the activations’ temporal structure specifically reflected that of the video, rather than regions whose activity was simply autocorrelated at a width similar to the video template’s diagonal.  To achieve this, we used a phase shift-based permutation procedure, wherein we circularly shifted the video’s topic trajectory by a random number of timepoints, computed the resulting ``null” video template, and re-ran the searchlight analysis, in full.  (For each of the 100 permutations, the same random shift was used for all participants).  We $z$-scored the observed (unshifted) result at each voxel against the distribution of permutation-derived ``null” results, and estimated a $p$-value by computing the proportion of shifted results that yielded larger values.  To create the map in Figure~\ref{fig:brainz}A, we thresholded out any voxels whose similarity to the unshifted video’s structure fell below the 95\textsuperscript{th} percentile of the permutation-derived similarity results.

We used an analogous procedure to identify which voxels' responses reflected the recall templates.  For each participant, we correlated the upper triangle of the correlation matrix for each cube of voxels with their (time warped) recall correlation matrix.  As in the video template analysis this yielded a voxelwise map of correlation coefficients per participant.  However, whereas the video analysis compared every participant's responses to the same template, here the recall templates were unique for each participant.  As in the analysis described above, we $t$-scored the (Fisher $z$-transformed) voxelwise correlations, and used the same permutation procedure we developed for the video responses to ensure specificity to the recall timeseries and assign significance values.  To create the map in Figure~\ref{fig:brainz}B we again thresholded out any voxels whose correspondence values fell below the 95\textsuperscript{th} percentile of the permutation-derived null distribution.

\bibliography{../../CDL-bibliography/memlab}
% \bibliography{CDL-bibliography/memlab}

>>>>

\subsubsection*{Visualizing the video and recall topic trajectories}
We used the UMAP algorithm~\citep{McInEtal18} to project the 100-dimensional topic space onto a two-dimensional space for visualization (Figs.~\ref{fig:trajectory}, \ref{fig:topics}).  Importantly, to ensure that all of the trajectories were projected onto the \textit{same} lower dimensional space, we computed the low-dimensional embedding on a ``stacked" matrix created by vertically concatenating the events-by-topics topic proportions matrices for the video, across-participants average recall and all 17 individual participants' recalls.  We then divided the rows of the result (a total-number-of-events by two matrix) back into separate matrices for the video topic trajectory and the trajectories for each participant's recalls (Fig.~\ref{fig:trajectory}).  This general approach for discovering a shared low-dimensional embedding for a collections of high-dimensional observations follows \cite{HeusEtal18a}.  Note: for further details on how we created this low-dimensional embedding space, see \textit{Supporting Information}.
% ------ NOTE: ------
% need to add info about embedding optimization
% - Paxton

\subsubsection*{Estimating the consistency of flow through topic space across participants}
In Figure~\ref{fig:trajectory}B, we present an analysis aimed at characterizing locations in topic space that different participants move through in a consistent way (via their recall topic trajectories).  The two-dimensional topic space used in our visualizations (Fig.~\ref{fig:trajectory}) comprised a 60 x 60 (arbitrary units) square.  We tiled this space with a 50 x 50 grid of evenly spaced vertices, and defined a circular area centered on each vertex whose radius was two times the distance between adjacent vertices (i.e., 2.4 units).  For each vertex, we examined the set of line segments formed by connecting each pair successively recalled events, across all participants, that passed through this circle.  We computed the distribution of angles formed by those segments and the $x$-axis, and used a Rayleigh test to determine whether the distribution of angles was reliably ``peaked'' (i.e., consistent across all transitions that passed through that local portion of topic space).  To create Figure~\ref{fig:trajectory}B we drew an arrow originating from each grid vertex, pointing in the direction of the average angle formed by the line segments that passed within its circular radius.  We set the arrow lengths to be inversely proportional to the $p$-values of the Rayleigh tests at each vertex.  Specifically, for each vertex we converted all of the angles of segments that passed within 2.4 units to unit vectors, and we set the arrow lengths at each vertex proportional to the length of the (circular) mean vector.  We also indicated any significant results ($p < 0.05$, corrected using the Benjamani-Hochberg procedure) by coloring the arrows in blue (darker blue denotes a lower $p$-value, i.e., a longer mean vector); all tests with $p \geq 0.05$ are displayed in gray and given a lower opacity value.

\subsection*{Searchlight fMRI analyses}
In Figure~\ref{fig:brainz}, we present two analyses aimed at identifying brain regions whose responses (as participants viewed the video) exhibited a particular temporal structure.  We developed a searchlight analysis wherein we constructed a 5 x 5 x 5 cube of voxels (following \citealp{ChenEtal17}) centered on each voxel in the brain, and for each of these cubes, computed the temporal correlation matrix of the voxel responses during video viewing.  Specifically, for each of the 1976 volumes collected during video viewing, we correlated the activity patterns in the given cube with the activity patterns (in the same cube) collected during every other timepoint.  This yielded a 1976 by 1976 correlation matrix for each cube.  Note: participant 5's scan ended 75s early, and in \citealp{ChenEtal17}'s publicly released dataset, their scan data was padded to match the length of the other participants'.  For our searchlight analyses, we removed this padded data (i.e., the last 50 TRs), resulting in a 1925 by 1925 correlation matrix for each cube in participant 5's brain.

Next, we constructed a series of ``template" matrices: the first reflecting the timecourse of video's topic trajectory, and the others reflecting that of each participant's recall topic trajectory.  To construct the video template, we computed the correlations between the topic proportions estimated for every pair of TRs (prior to segmenting the trajectory into discrete events; i.e., the correlation matrix shown in Figs.~\ref{fig:model}B and \ref{fig:brainz}A).  We constructed similar temporal correlation matrices for each participant's recall topic trajectory (Figs.~\ref{fig:model}D, \corrmats).  However, to correct for length differences and potential non-linear transformations between viewing time and recall time, we first used dynamic time warping~\citep{BernClif94} to temporally align participants' recall topic trajectories with the video topic trajectory.  An example correlation matrix before and after warping is shown in Fig.~\ref{fig:brainz}B.  This yielded a 1976 by 1976 correlation matrix for the video template and for each participant's recall template.

The temporal structure of the video's content (as described by our model) is captured in the block-diagonal structure of the video's temporal correlation matrix (e.g., Figs.~\ref{fig:model}B, \ref{fig:brainz}A), with time periods of thematic stability represented as dark blocks of varying sizes.  Inspecting the video correlation matrix suggests that the video's semantic content is highly temporally specific (i.e., the correlations between topic vectors from distant timepoints are almost entirely near-zero).  By contrast, the activity patterns of individual (cubes of) voxels can encode relatively limited information on their own, and their activity frequently contributes to multiple separate functions \citep{FreeEtal01, SigmDeha08, CharKoec10, RishEtal13}.  By nature, these two attributes give rise to similarities in activity across large timescales that may not necessarily reflect a single task.  To enable a more sensitive analysis of brain regions whose shifts in activity patterns mirrored shifts in the semantic content of the video or recalls, we restricted the temporal correlations we considered to timescale of semantic information captured by our model.  Specifically, we isolated the upper triangle of the video correlation matrix and created a ``proximal correlation mask" that included only diagonals from the upper triangle of the video correlation matrix up to the first that contained no positive correlations.  Applying this mask to the full video correlation matrix was analogous to excluding diagonals beyond the corner of the largest diagonal block.  In other words, the timescale of temporal correlations we considered corresponded to the longest period of thematic stability in the video, and by extension the longest expected period of thematic stability in participants' recalls and the longest period of stability we might expect to see in voxel activity arising from processing or encoding video content.  Figure \ref{fig:brainz} shows this proximal correlation mask applied to the temporal correlation matrices for the video, an example participant's (warped) recall, and an example cube of voxels from our searchlight analyses.

To determine which (cubes of) voxel responses matched the video template, we correlated the proximal diagonals from the upper triangle of the voxel correlation matrix for each cube with the proximal diagonals from video template matrix~\citep{KrieEtal08b}.  This yielded, for each participant, a voxelwise map of correlation values.  We then performed a one-sample $t$-test on the distribution of (Fisher $z$-transformed) correlations at each voxel, across participants.  This resulted in a value for each voxel (cube), describing how reliably its timecourse mirrored that of the video.

We further sought to ensure that our analysis identified regions where the activations’ temporal structure specifically reflected that of the video, rather than regions whose activity was simply autocorrelated at a width similar to the video template’s diagonal.  To achieve this, we used a phase shift-based permutation procedure, wherein we circularly shifted the video’s topic trajectory by a random number of timepoints, computed the resulting ``null” video template, and re-ran the searchlight analysis, in full.  (For each of the 100 permutations, the same random shift was used for all participants).  We $z$-scored the observed (unshifted) result at each voxel against the distribution of permutation-derived ``null” results, and estimated a $p$-value by computing the proportion of shifted results that yielded larger values.  To create the map in Figure~\ref{fig:brainz}C, we thresholded out any voxels whose similarity to the unshifted video’s structure fell below the 95\textsuperscript{th} percentile of the permutation-derived similarity results.

We used an analogous procedure to identify which voxels' responses reflected the recall templates.  For each participant, we correlated the proximal diagonals from the upper triangle of the correlation matrix for each cube of voxels with the proximal diagonals from the upper triangle of their (time-warped) recall correlation matrix.  As in the video template analysis, this yielded a voxelwise map of correlation coefficients per participant.  However, whereas the video analysis compared every participant's responses to the same template, here the recall templates were unique for each participant.  As in the analysis described above, we $t$-scored the (Fisher $z$-transformed) voxelwise correlations, and used the same permutation procedure we developed for the video responses to ensure specificity to the recall timeseries and assign significance values.  To create the map in Figure~\ref{fig:brainz}D we again thresholded out any voxels whose correspondence values fell below the 95\textsuperscript{th} percentile of the permutation-derived null distribution.

%\subsection*{Neurosynth decoding analyses}
%To gain further insights into the results from our video searchlight analysis, we constructed a voxelwise map of $p$-values by assigning each voxel the proportion of phase-shifted video trajectories to which its corresponding cube's timeseries was more highly correlated than the timeseries of the ``real" video.  This $p$-maps described the sensitivity of each voxel (cube) to the specific temporal structure of the video.  We then passed this $p$-map to \texttt{Neurosynth}.  Neurosynth does cool stuff.

% \bibliography{../../CDL-bibliography/memlab}
\bibliography{CDL-bibliography/memlab}
